---
title: "Bioinformatics Pipeline"
author: "Quynh-Chi Phan and Jennifer Zhang"
date: "2025-07-26"
output:
  pdf_document: default
---

## Install and load the following packages

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c("phyloseq", "microbiome", "ComplexHeatmap"), update = FALSE)
```

```{r}
library(tidyverse)
library(phyloseq)
library(vegan)
library(DESeq2)
library(ANCOMBC)
library(ComplexHeatmap)
library(microbiome)
library(knitr)
library(dplyr)
library(devtools)
library(data.table)
library(ggplot2)
```

## Load and clean the data

-   Imports the metadata, feature table, and taxonomy
-   Cleans up taxonomy files

All files were previously generated in QIIME2.

```{r Load and clean data}
otu <- read.table(file = "table.tsv", sep = "\t", header = T, row.names = 1, 
                  skip = 1, comment.char = "")

metadata <- read.table(file = "metadata.tsv", sep = "\t", header = T, row.names = 1)

taxonomy <- read.table(file = "taxonomy.tsv", sep = "\t", header = T ,row.names = 1)

tax <- taxonomy %>%
  select(Taxon) %>% 
  separate(Taxon, c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"), "; ")

tax.clean <- data.frame(row.names = row.names(tax),
                        Domain = str_replace(tax[,1], "d__",""),
                        Phylum = str_replace(tax[,2], "p__",""),
                        Class = str_replace(tax[,3], "c__",""),
                        Order = str_replace(tax[,4], "o__",""),
                        Family = str_replace(tax[,5], "f__",""),
                        Genus = str_replace(tax[,6], "g__",""),
                        Species = str_replace(tax[,7], "s__",""),
                        stringsAsFactors = FALSE)

tax.clean[is.na(tax.clean)] <- ""
tax.clean[tax.clean=="__"] <- ""

for (i in 1:nrow(tax.clean)){
  if (tax.clean[i,7] != ""){
    tax.clean$Species[i] <- paste(tax.clean$Genus[i], tax.clean$Species[i], sep = " ")
  } else if (tax.clean[i,2] == ""){
    kingdom <- paste("Unclassified", tax.clean[i,1], sep = " ")
    tax.clean[i, 2:7] <- domain
  } else if (tax.clean[i,3] == ""){
    phylum <- paste("Unclassified", tax.clean[i,2], sep = " ")
    tax.clean[i, 3:7] <- phylum
  } else if (tax.clean[i,4] == ""){
    class <- paste("Unclassified", tax.clean[i,3], sep = " ")
    tax.clean[i, 4:7] <- class
  } else if (tax.clean[i,5] == ""){
    order <- paste("Unclassified", tax.clean[i,4], sep = " ")
    tax.clean[i, 5:7] <- order
  } else if (tax.clean[i,6] == ""){
    family <- paste("Unclassified", tax.clean[i,5], sep = " ")
    tax.clean[i, 6:7] <- family
  } else if (tax.clean[i,7] == ""){
    tax.clean$Species[i] <- paste("Unclassified ",tax.clean$Genus[i], sep = " ")
  }
}
```

## Create complete Phyloseq object and filter out mitochondria and chloroplast.

```{r Create ps object}
OTU = otu_table(as.matrix(otu), taxa_are_rows = TRUE)
TAX = tax_table(as.matrix(tax.clean))
SAMPLE <- sample_data(metadata)
TREE = read_tree("tree.nwk")

# Merge data and save into PhyloSeq object 'ps'
# Filter out Eukaryotes, Archaea, mitochondria, and chloroplasts

#ps_all <- phyloseq(OTU, TAX, SAMPLE, TREE) %>%
ps_all <- phyloseq(OTU, TAX, SAMPLE) %>% ## DEBUG TREE
  subset_taxa(
    Domain == "Bacteria" &
    Family  != "mitochondria" &
    Class   != "Chloroplast"
  )

ps.comp = microbiome::transform(ps_all, "compositional")
```

## Subset ps_all and ps.comp based on year and location

```{r Subset year/location}
ps.filter_2023 <- subset_samples(ps_all, Year == 2023)
ps.filter_2024 <- subset_samples(ps_all, Year == 2024)
ps.filter_ud <- subset_samples(ps_all, Location != "MusselZone")
ps.filter_u <- subset_samples(ps_all, Location == "Upstream")
ps.filter_d <- subset_samples(ps_all, Location == "Downstream")
ps.filter_m <- subset_samples(ps_all, Location == "MusselZone")

ps.comp_2023 <- subset_samples(ps.comp, Year == 2023)
ps.comp_2024 <- subset_samples(ps.comp, Year == 2024)
ps.comp_ud <- subset_samples(ps.comp, Location != "MusselZone")
ps.comp_u <- subset_samples(ps.comp, Location == "Upstream")
ps.comp_d <- subset_samples(ps.comp, Location == "Downstream")
ps.comp_m <- subset_samples(ps.comp, Location == "MusselZone")
```

# ALPHA DIVERSITY

## Alpha diversity and pairwise comparison for 2023

```{r Alpha diversity 2023}
# Update factor levels in sample metadata
sample_data(ps.filter_2023)$Location <- factor(sample_data(ps.filter_2023)$Location,
                                        levels = c("Upstream", "MusselZone", "Downstream"))

# Creates boxplots depicting the observed and Shannon diversity between locations in 2023
plot_richness(ps.filter_2023, x="Location", measures=c("Observed", "Shannon")) +
              geom_boxplot(aes(fill = Location)) +
              scale_fill_brewer(palette = "Set2") +
              theme_classic() +
              theme(strip.background = element_blank(), axis.text.x.bottom = element_text(angle = -90)) +
              labs(x = "Location", 
                   y = "Diversity Measure")

rich = estimate_richness(ps.filter_2023, measures = c("Observed", "Shannon"))

rich$Location <- sample_data(ps.filter_2023)$Location
rich_summary <- rich %>%
                group_by(Location) %>%
                summarise(
                  Observed_Mean = mean(Observed),
                  Observed_SD   = sd(Observed),
                  Shannon_Mean  = mean(Shannon),
                  Shannon_SD    = sd(Shannon)
                ) %>%
                as.data.frame()
rich_summary

# Wilcoxon test based on observed diversity
wilcox.observed <- pairwise.wilcox.test(rich$Shannon, sample_data(ps.filter_2023)$Location, p.adjust.method = "BH")

# Format Wilcoxon test table
tab.observed <- wilcox.observed$p.value %>%
                as.data.frame() %>%
                rownames_to_column(var = "Group1") %>%
                pivot_longer(cols = -Group1, names_to = "Group2", values_to = "Adjusted_p_value") %>%
                filter(!is.na(Adjusted_p_value)) %>%
                mutate(Adjusted_p_value = signif(Adjusted_p_value, 3)) %>%
                arrange(Adjusted_p_value)

# View Wilcoxon test result for observed diversity
tab.observed

# Wilcoxon test based on Shannon diversity
wilcox.shannon <- pairwise.wilcox.test(rich$Shannon, sample_data(ps.filter_2023)$Location, p.adjust.method = "BH")

# Format Wilcoxon test table
tab.shannon <- wilcox.shannon$p.value %>%
               as.data.frame() %>%
               rownames_to_column(var = "Group1") %>%
               pivot_longer(cols = -Group1, names_to = "Group2", values_to = "Adjusted_p_value") %>%
               filter(!is.na(Adjusted_p_value)) %>%
               mutate(Adjusted_p_value = signif(Adjusted_p_value, 3)) %>%
               arrange(Adjusted_p_value)

# View Wilcoxon test result for Shannon diversity
tab.shannon
```

## Alpha diversity and pairwise comparison for 2024

```{r Alpha diversity 2024}
# Update factor levels in sample metadata
sample_data(ps.filter_2024)$Location <- factor(sample_data(ps.filter_2024)$Location,
                                        levels = c("Upstream", "MusselZone", "Downstream"))

# Creates boxplots depicting the observed and Shannon diversity between locations in 2024
plot_richness(ps.filter_2024, x="Location", measures=c("Observed", "Shannon")) +
              geom_boxplot(aes(fill = Location)) +
              scale_fill_brewer(palette = "Set2") +
              theme_classic() +
              theme(strip.background = element_blank(), axis.text.x.bottom = element_text(angle = -90)) +
              labs(x = "Location", 
                   y = "Diversity Measure")
              theme_classic() +
              theme(strip.background = element_blank(), axis.text.x.bottom = element_text(angle = -90))

rich = estimate_richness(ps.filter_2024, measures = c("Observed", "Shannon"))

rich$Location <- sample_data(ps.filter_2024)$Location
rich_summary <- rich %>%
                group_by(Location) %>%
                summarise(
                  Observed_Mean = mean(Observed),
                  Observed_SD   = sd(Observed),
                  Shannon_Mean  = mean(Shannon),
                  Shannon_SD    = sd(Shannon)
                ) %>%
                as.data.frame()
rich_summary

# Wilcoxon test based on observed diversity
wilcox.observed <- pairwise.wilcox.test(rich$Shannon, sample_data(ps.filter_2024)$Location, p.adjust.method = "BH")

# Format Wilcoxon test table
tab.observed <- wilcox.observed$p.value %>%
                as.data.frame() %>%
                rownames_to_column(var = "Group1") %>%
                pivot_longer(cols = -Group1, names_to = "Group2", values_to = "Adjusted_p_value") %>%
                filter(!is.na(Adjusted_p_value)) %>%
                mutate(Adjusted_p_value = signif(Adjusted_p_value, 3)) %>%
                arrange(Adjusted_p_value)

# View Wilcoxon test result for observed diversity
tab.observed

# Wilcoxon test based on Shannon diversity
wilcox.shannon <- pairwise.wilcox.test(rich$Shannon, sample_data(ps.filter_2024)$Location, p.adjust.method = "BH")

# Format Wilcoxon test table
tab.shannon <- wilcox.shannon$p.value %>%
               as.data.frame() %>%
               rownames_to_column(var = "Group1") %>%
               pivot_longer(cols = -Group1, names_to = "Group2", values_to = "Adjusted_p_value") %>%
               filter(!is.na(Adjusted_p_value)) %>%
               mutate(Adjusted_p_value = signif(Adjusted_p_value, 3)) %>%
               arrange(Adjusted_p_value)

# View Wilcoxon test result for Shannon diversity
tab.shannon
```

## Alpha diversity and pairwise comparison for all years

```{r Alpha diversity all years}
# Update factor levels in sample metadata
sample_data(ps_all)$LocYear <- factor(sample_data(ps_all)$LocYear,
                                      levels = c("Upstream_2023", "Upstream_2024", "MusselZone_2023", "MusselZone_2024", "Downstream_2023", "Downstream_2024"))


# Creates boxplots depicting the observed and Shannon values between locations and years
plot_richness(ps_all, x="LocYear", measures=c("Observed", "Shannon")) +
              geom_boxplot(aes(fill = LocYear)) +
              scale_fill_brewer(palette = "Set2") +
              theme_classic() +
              theme(strip.background = element_blank(), axis.text.x.bottom = element_text(angle = -90))

rich_all = estimate_richness(ps_all, measures = c("Observed", "Shannon"))
rich$LocYear <- sample_data(ps_all)$LocYear
rich_summary <- rich %>%
                group_by(LocYear) %>%
                summarise(
                  Observed_Mean = mean(Observed),
                  Observed_SD   = sd(Observed),
                  Shannon_Mean  = mean(Shannon),
                  Shannon_SD    = sd(Shannon)
                ) %>%
                as.data.frame()
rich_summary

# Wilcoxon test based on Shannon Diversity for 2023 and 2024
wilcox.shannon <- pairwise.wilcox.test(rich_all$Shannon, sample_data(ps_all)$LocYear, p.adjust.method = "BH")

# Format Wilcoxon test table
tab.shannon <- wilcox.shannon$p.value %>%
  as.data.frame() %>%
  rownames_to_column(var = "Group1") %>%
  pivot_longer(cols = -Group1, names_to = "Group2", values_to = "Adjusted_p_value") %>%
  filter(!is.na(Adjusted_p_value)) %>%
  mutate(Adjusted_p_value = signif(Adjusted_p_value, 3)) %>%
  arrange(Adjusted_p_value)

# View Wilcoxon test result for Shannon diversity
tab.shannon
```

## Alpha diversity and pairwise comparison between years for upstream zone

```{r Alpha diversity upstream}
# Creates boxplots depicting the observed and Shannon diversity between years
plot_richness(ps.filter_u, x="Year", measures=c("Observed", "Shannon")) +
              geom_boxplot(aes(fill = LocYear)) +
              scale_fill_brewer(palette = "Set2") +
              theme_classic() +
              theme(strip.background = element_blank(), axis.text.x.bottom = element_text(angle = -90))

rich_all = estimate_richness(ps.filter_u, measures = c("Observed", "Shannon"))

# Wilcoxon test based on Shannon Diversity for 2023 and 2024
wilcox.shannon <- pairwise.wilcox.test(rich_all$Shannon, sample_data(ps.filter_u)$Year, p.adjust.method = "BH")

# Format Wilcoxon test table
tab.shannon <- wilcox.shannon$p.value %>%
               as.data.frame() %>%
               rownames_to_column(var = "Group1") %>%
               pivot_longer(cols = -Group1, names_to = "Group2", values_to = "Adjusted_p_value") %>%
               filter(!is.na(Adjusted_p_value)) %>%
               mutate(Adjusted_p_value = signif(Adjusted_p_value, 3)) %>%
               arrange(Adjusted_p_value)

# View Wilcoxon test result for Shannon diversity
tab.shannon
```

## Alpha diversity and pairwise comparison between years for mussel zone

```{r Alpha diversity mussel}
# Creates boxplots depicting the observed and Shannon diversity between years
plot_richness(ps.filter_m, x="Year", measures=c("Observed", "Shannon")) +
              geom_boxplot(aes(fill = LocYear)) +
              scale_fill_brewer(palette = "Set2") +
              theme_classic() +
              theme(strip.background = element_blank(), axis.text.x.bottom = element_text(angle = -90))

rich_all = estimate_richness(ps.filter_m, measures = c("Observed", "Shannon"))

# Wilcoxon test based on Shannon Diversity for 2023 and 2024
wilcox.shannon <- pairwise.wilcox.test(rich_all$Shannon, sample_data(ps.filter_m)$Year, p.adjust.method = "BH")

# Format Wilcoxon test table
tab.shannon <- wilcox.shannon$p.value %>%
               as.data.frame() %>%
               rownames_to_column(var = "Group1") %>%
               pivot_longer(cols = -Group1, names_to = "Group2", values_to = "Adjusted_p_value") %>%
               filter(!is.na(Adjusted_p_value)) %>%
               mutate(Adjusted_p_value = signif(Adjusted_p_value, 3)) %>%
               arrange(Adjusted_p_value)

# View Wilcoxon test result for Shannon diversity
tab.shannon
```

## Alpha diversity and pairwise comparison between years for downstream zone

```{r Alpha diversity mussel}
# Creates boxplots depicting the observed and Shannon diversity between years
plot_richness(ps.filter_d, x="Year", measures=c("Observed", "Shannon")) +
              geom_boxplot(aes(fill = LocYear)) +
              scale_fill_brewer(palette = "Set2") +
              theme_classic() +
              theme(strip.background = element_blank(), axis.text.x.bottom = element_text(angle = -90))

rich_all = estimate_richness(ps.filter_d, measures = c("Observed", "Shannon"))

# Wilcoxon test based on Shannon Diversity for 2023 and 2024
wilcox.shannon <- pairwise.wilcox.test(rich_all$Shannon, sample_data(ps.filter_d)$Year, p.adjust.method = "BH")

# Format Wilcoxon test table
tab.shannon <- wilcox.shannon$p.value %>%
               as.data.frame() %>%
               rownames_to_column(var = "Group1") %>%
               pivot_longer(cols = -Group1, names_to = "Group2", values_to = "Adjusted_p_value") %>%
               filter(!is.na(Adjusted_p_value)) %>%
               mutate(Adjusted_p_value = signif(Adjusted_p_value, 3)) %>%
               arrange(Adjusted_p_value)

# View Wilcoxon test result for Shannon diversity
tab.shannon
```

# BETA DIVERSITY

## PCoA using Bray Curtis distances for 2023

```{r PCoA 2023}
dist = phyloseq::distance(ps.filter_2023, method="bray")
ordination = ordinate(ps.filter_2023, method="PCoA", distance=dist)
plot_ordination(ps.filter_2023, ordination, color="Location", title="2023 PCoA by Location") + 
                theme_classic() +
                theme(strip.background = element_blank())
cbn <- combn(x=unique(metadata$Location), m = 2)
p <- c()

for(i in 1:ncol(cbn)){
  ps.subs <- subset_samples(ps.filter_2023, Location %in% cbn[,i])
  metadata_sub <- data.frame(sample_data(ps.subs))
  permanova_pairwise <- adonis2(phyloseq::distance(ps.subs, method = "bray") ~ Location, 
                                data = metadata_sub)
  p <- c(p, permanova_pairwise$`Pr(>F)`[1])
}
p.adj <- p.adjust(p, method = "BH")
p.table.permanova <- cbind.data.frame(t(cbn), p=p, p.adj=p.adj)
p.table.permanova
```

## PCoA using Bray Curtis distances for 2024

```{r PCoA 2024}
dist = phyloseq::distance(ps.filter_2024, method="bray")
ordination = ordinate(ps.filter_2024, method="PCoA", distance=dist)
plot_ordination(ps.filter_2024, ordination, color="Location", title="2024 PCoA by Location") +
                theme_classic() +
                theme(strip.background = element_blank())

cbn <- combn(x=unique(metadata$Location), m = 2)
p <- c()

for(i in 1:ncol(cbn)){
  ps.subs <- subset_samples(ps.filter_2024, Location %in% cbn[,i])
  metadata_sub <- data.frame(sample_data(ps.subs))
  permanova_pairwise <- adonis2(phyloseq::distance(ps.subs, method = "bray") ~ Location, 
                                data = metadata_sub)
  p <- c(p, permanova_pairwise$`Pr(>F)`[1])
}
p.adj <- p.adjust(p, method = "BH")
p.table.permanova <- cbind.data.frame(t(cbn), p=p, p.adj=p.adj)
p.table.permanova
```

## PCOA for both years

```{r PCoA for both years}
dist = phyloseq::distance(ps.filter_all, method="bray")
ordination = ordinate(ps.filter_all, method="PCoA", distance=dist)
plot_ordination(ps.filter_all, ordination, color="LocYear", title="Overall PCoA by Location and Year") +
                theme_classic() +
                theme(strip.background = element_blank())

cbn <- combn(x=unique(metadata$LocYear), m = 2)
p <- c()

for(i in 1:ncol(cbn)){
  ps.subs <- subset_samples(ps.filter_all, LocYear %in% cbn[,i])
  metadata_sub <- data.frame(sample_data(ps.subs))
  permanova_pairwise <- adonis2(phyloseq::distance(ps.subs, method = "bray") ~ LocYear, 
                               data = metadata_sub)
  p <- c(p, permanova_pairwise$`Pr(>F)`[1])
}
p.adj <- p.adjust(p, method = "BH")
p.table.permanova <- cbind.data.frame(t(cbn), p=p, p.adj=p.adj)
p.table.permanova
```

# ANCOM-BC2 ANALYSIS WITH HOLM METHOD

## ANCOM-BC2 between upstream and downstream for both 2023 and 2024
https://www.bioconductor.org/packages/release/bioc/vignettes/ANCOMBC/inst/doc/ANCOMBC2.html
https://www.bioconductor.org/packages/devel/bioc/manuals/ANCOMBC/man/ANCOMBC.pdf
```{r ANCOM for both years}
set.seed(123)

sample_data(ps.filter_ud)$Location <- as.factor(sample_data(ps.filter_ud)$Location)

pseq_perm = ps.filter_ud
metadata_perm = microbiome::meta(pseq_perm)
metadata_perm$Location = sample(metadata_perm$Location)
phyloseq::sample_data(pseq_perm) = metadata_perm

# Run ancombc2 function
output.family = ancombc2(data = pseq_perm, tax_level = "Family",
                         fix_formula = "Location", rand_formula = NULL,
                         group = "Location", struc_zero = TRUE, neg_lb = TRUE,
                         n_cl = 4, verbose = TRUE,
                         global = TRUE, pairwise = TRUE, dunnet = TRUE, trend = TRUE,
                         lme_control = lme4::lmerControl())

# Primary analysis
res_family = output.family$res

# Results for location
df_location = res_family %>%
              dplyr::select(taxon, contains("Location"))
df_fig_location = df_location %>%
                  dplyr::filter(p_LocationUpstream < 0.05) %>%
                  dplyr::mutate(lfc = round(lfc_LocationUpstream, 2),
                                color = ifelse(passed_ss_LocationUpstream == 1, "black", "white")) %>%
                                dplyr::arrange(taxon)
df_fig_location

lo = floor(min(df_fig_location$lfc))
up = ceiling(max(df_fig_location$lfc))
mid = (lo + up) / 2
fig_location = df_fig_location %>%
 ggplot(aes(x = "Downstream Samples", y = taxon, fill = lfc)) +
 geom_tile(color = "black") +
 scale_fill_gradient2(low = "lightblue", high = "navy", mid = "blue",
 na.value = "white", midpoint = mid, limit = c(lo, up),
name = NULL) +
 geom_text(aes(label = lfc, color = color), size = 4) +
 scale_color_identity(guide = "none") +
 labs(x = NULL, y = NULL, title = "Differentually Abundant Families as Compared to Upstream Samples") +
 theme_minimal() +
 theme(plot.title = element_text(hjust = 0.5))
fig_location
```

## Between upstream and downstream in 2023.

```{r}
sample_data(ps.filter_2023)$donor_status <- as.factor(sample_data(ps.filter_2023)$Location)
set.seed(123)
pseq_perm = ps.filter_2023 #make sure to choose either your
metadata_perm = microbiome::meta(pseq_perm)
metadata_perm$Location = sample(metadata_perm$Location)
phyloseq::sample_data(pseq_perm) = metadata_perm

set.seed(123)
# It should be noted that we have set the number of bootstrap samples (B) equal
# to 10 in the 'trend_control' function for computational expediency.
# However, it is recommended that users utilize the default value of B,
# which is 100, or larger values for optimal performance.
output.family = ancombc2(data = pseq_perm, tax_level = "Family",
 fix_formula = "Location", rand_formula = NULL,
 p_adj_method = "holm", pseudo_sens = TRUE,
 prv_cut = 0.10, lib_cut = 1000, s0_perc = 0.05,
 group = "Location", struc_zero = TRUE, neg_lb = TRUE,
 alpha = 0.05, n_cl = 2, verbose = TRUE,
 global = TRUE, pairwise = TRUE, dunnet = TRUE, trend = TRUE,
 iter_control = list(tol = 1e-2, max_iter = 20,
 verbose = TRUE),
 em_control = list(tol = 1e-5, max_iter = 100),
 lme_control = lme4::lmerControl(),
 mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
 trend_control = list(contrast = list(matrix(c(1, 0, -1, 1),
 nrow = 2,
byrow = TRUE),
 matrix(c(-1, 0, 1, -1),
nrow = 2,
byrow = TRUE),
 matrix(c(1, 0, 1, -1),
nrow = 2,
byrow = TRUE)),
 node = list(2, 2, 1),
solver = "ECOS",
B = 100))

res_family = output.family$res
df_location = res_family %>%
 dplyr::select(taxon, contains("Location"))
df_fig_location = df_location %>%
 dplyr::filter(p_LocationUpstream < 0.05) %>%
 #dplyr::filter(diff_donor_statusPD == 1) %>% ##Nothing was DA, so switched to p-value
 dplyr::mutate(lfc = round(lfc_LocationUpstream, 2),
 color = ifelse(passed_ss_LocationUpstream == 1, "black", "white")) %>%
 dplyr::arrange(taxon)
df_fig_location

lo = floor(min(df_fig_location$lfc))
up = ceiling(max(df_fig_location$lfc))
mid = (lo + up) / 2
fig_location = df_fig_location %>%
 ggplot(aes(x = "Downstream Samples", y = taxon, fill = lfc)) +
 geom_tile(color = "black") +
 scale_fill_gradient2(low = "lightblue", high = "navy", mid = "blue",
 na.value = "white", midpoint = mid, limit = c(lo, up),
name = NULL) +
 geom_text(aes(label = lfc, color = color), size = 4) +
 scale_color_identity(guide = "none") +
 labs(x = NULL, y = NULL, title = "Differentually Abundant Families as Compared to Upstream Samples") +
 theme_minimal() +
 theme(plot.title = element_text(hjust = 0.5))
fig_location
```

Between upstream and downstream for 2024.

```{r}
sample_data(ps.filter_2024)$donor_status <- as.factor(sample_data(ps.filter_2024)$Location)
set.seed(123)
pseq_perm = ps.filter_2024 #make sure to choose either your
metadata_perm = microbiome::meta(pseq_perm)
metadata_perm$Location = sample(metadata_perm$Location)
phyloseq::sample_data(pseq_perm) = metadata_perm

set.seed(123)
# It should be noted that we have set the number of bootstrap samples (B) equal
# to 10 in the 'trend_control' function for computational expediency.
# However, it is recommended that users utilize the default value of B,
# which is 100, or larger values for optimal performance.
output.family = ancombc2(data = pseq_perm, tax_level = "Family",
 fix_formula = "Location", rand_formula = NULL,
 p_adj_method = "holm", pseudo_sens = TRUE,
 prv_cut = 0.10, lib_cut = 1000, s0_perc = 0.05,
 group = "Location", struc_zero = TRUE, neg_lb = TRUE,
 alpha = 0.05, n_cl = 2, verbose = TRUE,
 global = TRUE, pairwise = TRUE, dunnet = TRUE, trend = TRUE,
 iter_control = list(tol = 1e-2, max_iter = 20,
 verbose = TRUE),
 em_control = list(tol = 1e-5, max_iter = 100),
 lme_control = lme4::lmerControl(),
 mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
 trend_control = list(contrast = list(matrix(c(1, 0, -1, 1),
 nrow = 2,
byrow = TRUE),
 matrix(c(-1, 0, 1, -1),
nrow = 2,
byrow = TRUE),
 matrix(c(1, 0, 1, -1),
nrow = 2,
byrow = TRUE)),
 node = list(2, 2, 1),
solver = "ECOS",
B = 100))

res_family = output.family$res
df_location = res_family %>%
 dplyr::select(taxon, contains("Location"))
df_fig_location = df_location %>%
 dplyr::filter(p_LocationUpstream < 0.05) %>%
 #dplyr::filter(diff_donor_statusPD == 1) %>% ##Nothing was DA, so switched to p-value
 dplyr::mutate(lfc = round(lfc_LocationUpstream, 2),
 color = ifelse(passed_ss_LocationUpstream == 1, "black", "white")) %>%
 dplyr::arrange(taxon)
df_fig_location


```
